Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-18 doc/thesis/discussion.tex
More writing

>>>>

In retrospect, there are a few algorithmic improvements which could have been
made to \software{netabc}. Firstly, our implementation uses simple multinomial
sampling for the resampling step, as well as for generating the final sample
from the target distribution.  Several other sampling strategies have been
developed~\autocite{douc2005comparison}, and it is possible that the use of a
more sophisticated scheme might increase the algorithm's accuracy. Secondly,
the model parameters \gls{I}, \gls{N}, and \gls{m} are all integer-valued, but
we treat them as continuous-valued when performing \gls{SMC}. This is not
inherently wrong, in that it simply assigns equal probability density to the
interval $[i, i+1)$ for all integers $i$. However, computational effort could
be wasted if particles are moved around within these intervals. To improve
efficiency, alternative proposals could be used for discrete-valued particles.

### 2016-04-18 doc/thesis/intro.tex
More writing

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the the evolutionary relationships among a group of
organisms or species (generally, \defn{taxa})~\autocite{haeckel1866generelle}.
The \defn{tips} of a phylogeny, that is, the nodes without any descendants,
correspond to \defn{extant}, or observed, taxa, while the \defn{internal nodes}
correspond to their common ancestors. The edges or \defn{branches} of the
phylogeny connect ancestors to their descendants. Phylogenies may have a
\defn{root}, which is a node with no descendants distinguished as the most
recent common ancestor of all the extant
taxa~\autocite{harding1971probabilities}. When such a root exists, the tree is
referred to as being \defn{rooted}; otherwise, it is \defn{unrooted}. The
structural arrangement of nodes and edges in the tree is referred to as its
\defn{topology}~\autocite{cavalli1967phylogenetic}.

@@

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population. Like phylogenies, transmission trees
have tips, nodes, edges, and branch lengths. However, rather than recording an
evolutionary process (speciation), they record an epidemiological process
(transmission). The tips of a transmission tree represent infected hosts, while
internal nodes correspond to transmissions from one host to another.
Transmission trees generally have branch lengths in units of calendar time,
with branching times indicating times of transmission. The root of a
transmission tree corresponds to the initially infected patient who introduced
the epidemic into the network, also known as the \defn{index case}. The
internal nodes may be labelled with the donor of the transmission pair, if this
is known. The tips of the tree, rather than being fixed at the present day, are
placed at the time at which the individual was removed from the epidemic, such
as by death, recovery, isolation, behaviour change, or migration. Consequently,
the transmission tree may not be ultrametric, but may have tips located at
varying distances from the root. Such trees are said to have
\defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast to
the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
transmission tree is illustrated in \cref{fig:contactnet} (right).

Due to the internal nodes, each infected individual in an epidemic may appear
in the transmission tree more than once. This is different from the
transmission \emph{network}, in which each infected individual appears exactly
once, and edges are in one-to-one correspondence with
transmissions~\autocite{welch2011statistical, keeling2005networks}.
Transmission networks are discussed further in \cref{sec:contactnet}, and the
distinction between the two objects is illustrated in \cref{fig:contactnet}.
However, since transmission networks generally have no cycles (unless
re-infection occurs), they are trees in the graph theoretical sense, and hence
are sometimes also referred to as transmission
trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
``transmission tree'' for the objects depicted on the right side of
\cref{fig:contactnet}, following~\autocite[\eg][]{stadler2013uncovering}. The
term ``transmission network'' is taken to mean the subgraph of the contact
network along which transmissions occurred,
following~\autocite{welch2011statistical, keeling2005networks}.

@@

Since transmission trees are essentially a detailed record of an epidemic's
progress, they contain substantial epidemiological information. As a basic
example, the \gls{ltt} plot~\autocite{nee1992tempo}, which plots the number of
lineages in a phylogeny against time, can be used to quantify the incidence of
new infections over the course of an epidemic~\autocite{holmes1995revealing}.
Many more diverse epidemiological parameters have been investigated using
transmission trees, such as the degree of
clustering~\autocite{hughes2009molecular} and the effect of elevated
transmission risk in acute infection~\autocite{volz2012simple}. However, in all
but the most well-studied of epidemics, this is not possible to obtain through
traditional epidemiological methods~\autocite{welch2011statistical}. The time
and effort to conduct detailed interviews and contact tracing of a sufficient
number of infected individuals is usually prohibitive. Even when the resources
for such methods are available, patients may not always recall whom they
contacted and when, especially in the case of airborne transmission.
Consequently, the transmission tree must be estimated using other methods. Most
commonly, this is done by exploiting the relationship between transmission
trees and viral phylogenies~\autocite{volz2013viral}.

\subsection{Relationship between transmission trees and viral phylogenies}

In general, \defn{viral phylogenies} are simply phylogenetic trees relating
virus strains. In phylodynamics, we often consider \defn{inter-host}
phylogenies, which relate one viral genotype from each host in a population.
The crux of \defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact
that the epidemiological processes recorded in transmission trees, and the
evolutionary processes recorded in viral phylogenies, occur on similar time
scales for RNA viruses~\autocite{drummond2003measurably}. As a result, there is
a close relationship between the two types of tree. In particular, the
transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual from the
transmission tree causes the extinction of their viral lineage in the
phylogeny. Due to these relationships, the topology of the viral phylogeny is
often used as a proxy for the topology of the transmission tree. However, there
are several complications and caveats which must be kept in mind when
estimating the transmission tree in this manner.

@@

A final caveat is that the viral phylogeny itself is not known with certainty,
so it must also be estimated from genetic data. Phylogenetic inference is a
complex topic which we will not discuss in detail here (see e.g.
\autocite{nei2000molecular} for a full review). Most modern analyses use
model-based methods, which simultaneously estimate the phylogeny with branch
lengths and the parameters of a model of evolution. Although they usually work
well in practice, the estimated topology can vary based on the model used and,
in the case of Bayesian analysis, the priors.  In addition, intra-host viral
populations are genetically heterogeneous, so choosing a single representative
genotype per host is necessarily imprecise. One can use either the genotype of
a specific virion sampled from the host, or a synthetic genotype, such as a
consensus or reconstructed ancestral sequence.

@@

@@

\subsection{Applications of phylodynamics}
\label{subsec:phyloapp}

Phylodynamic methods have been used to investigate epidemiological parameters
such as transmission rate, recovery rate, and basic reproductive
number~\autocite{pybus2009evolutionary, volz2013viral}. These studies make
inferences about epidemiological processes from the genetic diversity of virus
populations, which is usually represented in the form of a phylogeny. The
majority of these employ a Bayesian \gls{MCMC} approach to infer parameters of
an epidemiological model whose likelihood can be calculated, most often some
variation of the birth-death~\autocite{kendall1948generalized} or
coalescent~\autocite{kingman1982coalescent} models.
\textcite{stadler2012estimating} develop a formula for the likelihood of a
phylogeny with heterochronous tips under the birth-death model, which has been
used to estimate the basic reproductive number of several viral
epidemics~\autocite{stadler2012estimating}. However, the birth-death model is
cannot tell us anything about population structure, as it assumes that every
individual becomes infected at the same rate. \textcite{volz2012complex} writes
down the likelihood of a heterochronous phylogeny under a coalescent model with
arbitrarily complex population dynamics. This opens the door to more complex
inferences about population structure, as the population can be partitioned
into compartments with different transmission and recovery rates, but still
assumes that each compartment is homogeneously mixed. In other words, the
coalescent model can tell us about the \emph{global} structure of a population,
such as whether there exists a high-risk subgroup, but not about the
\emph{local} structure, such as the average number of contacts each individual
has.

@@

@@

\subsection{Sequential Monte Carlo samplers}

@@

Both algorithms involve progression through a sequence of related
distributions. For \gls{SIS}, these distributions are lower-dimensional
marginals of the target distribution, while for the \gls{SMC} sampler, they are
of the same dimension and constitute a smooth progression from an initial to a
final distribution. In both cases, the neighbouring distributions in the
sequence are related to each other in some way, and we can take advantage of
that relationship to create a sequence of importance distributions alongside
the sequence of targets. In \gls{SIS}, the neighbouring marginals
$\pi(\vec{x_i})$ and $\pi(\vec{x_{i+1}})$ were related by the conditional
density $\pi(x_i \mid \vec{x_{i-1}})$, which we used to inform the importance
distribution. In \gls{SMC}, the relationship between subsequent distributions
is less explicit, but it is assumed that they are related closely enough that
an importance distribution for $\pi_i$ can be easily transformed into one for
$\pi_{i+1}$. In particular, the sequence of importance distributions $\eta_i$
is constructed as
\begin{align}
  \label{eq:impint}
  \eta_i(x) = \int \eta_{i-1}(\xi) K_i(\xi, x) \d\xi,
\end{align}
where $K_i$ is any Markov kernel and the integral is over the space on which the
$\pi_i$ are defined. The choice of $K_i$ should be based on the percieved
relationship between $\pi_{i-1}$ and $\pi_i$. \textcite{del2006sequential}
propose the use of a \gls{MCMC} kernel with equilibrium distribution $\pi_i$.
That is,
\[
  K_i(\xi, x) = \max\left(1, \frac{q(x, \xi)\pi_i(\xi)}{q(\xi, x)\pi_i(x)}\right),
\]
where $q(\xi, x)$ is a proposal function such as a Gaussian distribution
centered at $\xi$ (see \cref{subsec:mfit}).

Although this method of building up $\eta$ appears straightforward, the
drawback is that the importance distribution itself becomes intractible. In
particular, evaluating $\eta_d(x)$ involves a $d$-dimensional integral of the
type \cref{eq:impint}. As it is necessary to evaluate $\eta(x)$ pointwise to
perform \gls{IS}, this construction appears to have defeated the purpose of
providing an importance distribution for each $\pi_i$.
\textcite{del2006sequential} overcome this problem with two ``artificial''
objects. First, they propose the existence of \textit{backward} kernels
$L_{i-1}(x_i, x_{i-1})$ which perform the inverse operation as $K_i$, namely
transforming $\eta_i$ back into $\eta_{i-1}$. As we shall see, these backward
kernels do not need to be calculated analytically in terms of $K_i$. Second,
they define a sequence of $i$-dimensional target distributions
\[
  \tilde{\pi}_i(\vec{x_i}) = \frac{\gamma_i(x_i)}{Z_i} \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)
\]
which admit $\pi_i(\vec{x_i})$ as marginals.

@@

Ordinarily, Bayesian inference targets the posterior distribution $\Pr(\theta
\mid \vec{y})$. That is, in the Bayesian framework, model parameters with
higher posterior density are ``better'' in the sense that they offer a more
credible explanation for the observed data. Approximate Bayesian computation
offers an alternative metric for parameter credibility, namely the similarity
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, suppose we have a distance measure $\rho$ defined on the space of all
possible data our model could generate. \gls{ABC} aims to sample from the joint
posterior distribution of model parameters and simulated datasets $\vec{z}$
which are within some small distance $\varepsilon$ of the observed data
$\vec{y}$,
\[
  \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon) =
  \frac{\Pr(\theta) \Pr(\vec{z} \mid \theta) \I_{A_{\varepsilon, \vec{y}}} (\vec{z})}
  {\int_{A_{\varepsilon, \vec{y}} \times \Theta} \Pr(\theta) \Pr(\vec{z} \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, \vec{y}}$ is an $\varepsilon$-ball around $\vec{y}$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
see in the next section, this distribution can be sampled from exactly. The
word ``approximate'' derives from the assumption that, for a suitably chosen
distance $\rho$ and a small enough $\varepsilon$, the marginal in $\vec{z}$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon) \d\vec{z}
  \approx \Pr(\theta \mid \vec{y}).
\]
To emphasize that the distribution on the left only approximates the posterior,
we will refer to it as the \textit{\gls{ABC} target distribution}. Note that in
many formulations, the distance function $\rho$ is defined as $\rho(S(\cdot),
S(\cdot))$ where $S$ is a function which maps data points into a vector of
summary statistics. This can be useful if the data are high-dimensional or of a
complex type, but it is not strictly necessary. For instance, if the data are
numeric and of low dimension, the distance function may simply be the Euclidian
distance~\autocite{sisson2007sequential}. For more complex data,
\textcite{nakagome2013kernel} proposed the use of a kernel function (defined in
\cref{subsec:treeshape}), an approach they dubbed \defn{kernel-\gls{ABC}}.

@@

Algorithms for performing \gls{ABC} fall into one of three categories:
rejection, \gls{MCMC}, and \gls{SMC}. Rejection ABC is the simplest method, and
also the one which was first proposed~\autocite{rubin1984bayesianly,
tavare1997inferring}. The algorithm, outlined in \cref{alg:abcrej}, repeats the
following steps until a desired number of samples from the target distribution
are obtained. Parameter values $\theta$ are sampled according to the prior
distribution $\Pr(\theta)$. Then, a simulated dataset $\vec{z}$ is generated
from the model with the sampled parameter values. By definition, the
probability density of obtaining the particular dataset $\vec{z}$ is
$\Pr(\vec{z} \mid \theta)$. Finally, the parameters are sampled if the distance
of $\vec{z}$ from the observed data $\vec{y}$ is less than $\varepsilon$ - that
is, with probability $\I_{A_{\varepsilon, \vec{y}}}(\vec{z})$. Putting this all
together, the parameters $\theta$ are sampled with probability proportional to
\[
  \Pr(\theta) \Pr(\vec{z} \mid \theta) \I_{A_{\varepsilon, \vec{y}}}(\vec{z}),
\]
which is exactly the numerator of the \gls{ABC} target distribution. Thus,
$\theta$ represents an unbiased sample from the approximate posterior.

@@

The most recently developed class of algorithm for \gls{ABC} is
\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, 2009adaptive}. As with
\gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of an
existing Bayesian inference method, in this case \gls{SMC} (\cref{sec:smc}).

>>>>

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the the evolutionary relationships among a group of
organisms or species (generally, \defn{taxa})~\autocite{haeckel1866generelle}.
The \defn{tips} of a phylogeny, that is, the nodes without any descendants,
correspond to \defn{extant}, or observed, taxa. The \defn{internal nodes}
correspond to their (usually extinct) common ancestors. The edges or
\defn{branches} of the phylogeny connect ancestors to their descendants.
Phylogenies may have a \defn{root}, which is a node with no descendants
distinguished as the most recent common ancestor of all the extant
taxa~\autocite{harding1971probabilities}. When such a root exists, the tree is
referred to as being \defn{rooted}; otherwise, it is \defn{unrooted}. The
structural arrangement of nodes and edges in the tree is referred to as its
\defn{topology}~\autocite{cavalli1967phylogenetic}.

@@

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population. Like phylogenies, transmission trees
have tips, nodes, edges, and branch lengths. However, rather than recording an
evolutionary process (speciation), they record an epidemiological process
(transmission). The tips of a transmission tree represent the removal of
infected hosts, while internal nodes correspond to transmissions from one host
to another.Transmission trees generally have branch lengths in units of
calendar time, with branching times indicating times of transmission. The root
of a transmission tree corresponds to the initially infected patient who
introduced the epidemic into the network, also known as the \defn{index case}.
The internal nodes may be labelled with the donor of the transmission pair, if
this is known. The tips of the tree, rather than being fixed at the present
day, are placed at the time at which the individual was removed from the
epidemic, such as by death, recovery, isolation, behaviour change, or
migration. Consequently, the transmission tree may not be ultrametric, but may
have tips located at varying distances from the root. Such trees are said to
have \defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast
to the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
transmission tree is illustrated in \cref{fig:contactnet} (right).

Each infected individual in an epidemic may appear in the transmission tree
more than once. This is different from the transmission \emph{network}, in
which each infected individual appears exactly once, and edges are in
one-to-one correspondence with transmissions~\autocite{welch2011statistical,
keeling2005networks}. Transmission networks are discussed further in
\cref{sec:contactnet}, and the distinction between the two objects is
illustrated in \cref{fig:contactnet}. However, since transmission networks
generally have no cycles (unless re-infection occurs), they are trees in the
graph theoretical sense, and hence are sometimes also referred to as
transmission trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we
reserve the term ``transmission tree'' for the objects depicted on the right
side of \cref{fig:contactnet}, following \eg~\autocite{stadler2013uncovering}.
The term ``transmission network'' is taken to mean the subgraph of the contact
network along which transmissions occurred, following
\eg~\autocite{welch2011statistical, keeling2005networks}.

@@

Since transmission trees are essentially a detailed record of an epidemic's
progress, they contain substantial epidemiological information. As a basic
example, the \gls{ltt} plot~\autocite{nee1992tempo}, which plots the number of
lineages in a phylogeny against time, can be used to quantify the incidence of
new infections over the course of an epidemic~\autocite{holmes1995revealing}.
However, in all but the most well-studied of epidemics, transmission trees are
not possible to assemble through traditional epidemiological
methods~\autocite{welch2011statistical}. The time and effort to conduct
detailed interviews and contact tracing of a sufficient number of infected
individuals is usually prohibitive, and may be additionally be confounded by
misreporting~\autocite{eames2015six}. However, it turns out that for viral
epidemics, some of the epidemiological information contained in the
transmission tree leaves a mark on the viral genetic material circulating in
the population. A family of methods called
\defn{phylodynamics}~\autocite{grenfell2004unifying} addresses the challenge of
estimating epidemiological parameters from viral sequence
data~\autocite{volz2013viral}.

\subsection{Phylodynamics: linking evolution and epidemiology}

%Most modern analyses use model-based methods, which simultaneously
%estimate the phylogeny with branch lengths and the parameters of a model of
%evolution. Although they usually work well in practice, the estimated topology
%can vary based on the model used and, in the case of Bayesian analysis, the
%priors. In addition, intra-host viral populations are genetically
%heterogeneous, so choosing a single representative genotype per host is
%necessarily imprecise. One can use either the genotype of a specific virion
%sampled from the host, or a synthetic genotype, such as a consensus or
%reconstructed ancestral sequence.

In general, \defn{viral phylogenies} are simply phylogenetic trees relating
virus strains. \defn{Inter-host} phylogenies relate one representative viral
genotype from each host in a population. In this work, when we use the phrase
``viral phylogeny'', we are referring to an inter-host phylogeny unless
otherwise stated. Phylogenetic inference is a complex topic which we will not
discuss here (see \eg \autocite{nei2000molecular} for a full review). We assume
here that the viral phylogeny is known, a common assumption in the development
and application of phylodynamic methods.

The crux of phylodynamics is the fact that the epidemiological processes
recorded in transmission trees, and the evolutionary processes recorded in
viral phylogenies, occur on similar time scales for RNA
viruses~\autocite{drummond2003measurably}. Phylodynamic methods have been used
to investigate epidemiological parameters such as transmission rate, recovery
rate, and basic reproductive number~\autocite{pybus2009evolutionary,
volz2013viral}. The majority of phylodynamic studies attempt to infer the
parameters of an epidemiological model for which the likelihood of an observed
phylogeny can be calculated. Most often, this is some variation of the
birth-death~\autocite{kendall1948generalized, stadler2012estimating} or
coalescent~\autocite{kingman1982coalescent, volz2012complex} models.

Due to the relationship between the aforementioned processes, there is a close
relationship between viral phylogenies and transmission trees. In particular,
the transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual from the
transmission tree causes the extinction of their viral lineage in the
phylogeny. Due to these relationships, the topology of the viral phylogeny is
often used as a proxy for the topology of the transmission tree. However, there
are several complications and caveats which must be kept in mind when
estimating the transmission tree in this manner.

@@

@@

\textcite{poon2013mapping} developed an alternative pairwise measure which
applies the concept of a \defn{kernel function} to phylogenies. Kernel
functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
compare objects by mapping them into a feature space of very high, possibly
even infinite, dimension.

@@

@@

To fully describe \gls{SMC}, we will introduce some notation and terminology.
For a sequence $x_1, \ldots, x_d$, we will write $\vec{x_i}$ to mean the
partial sequence $x_1, \ldots, x_i$.

A \defn{Markov kernel} is the continuous analogue of the transition matrix in a
finite-state Markov model. For some spaces $X$ and $Y$, $K \maps X \times Y \to
[0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. A more precise definition can be found in
\textbf{TODO}. Note that a Markov kernels have nothing to do with the kernel
functions defined in \cref{subsec:treeshape}, other than sharing a name.
Unfortunately, the word ``kernel'' is ubiquitous in mathematics and refers to
many unrelated concepts.

@@

\subsection{Sequential Monte Carlo samplers}
\label{subsec:smcsamp}

@@

Both algorithms involve progression through a sequence of related
distributions. For \gls{SIS}, these distributions are lower-dimensional
marginals of the target distribution, while for the \gls{SMC} sampler, they are
of the same dimension and constitute a smooth progression from an initial to a
final distribution. In both cases, the neighbouring distributions in the
sequence are related to each other in some way, and we can take advantage of
that relationship to create a sequence of importance distributions alongside
the sequence of targets. In \gls{SIS}, the neighbouring marginals
$\pi(\vec{x_i})$ and $\pi(\vec{x_{i+1}})$ were related by the conditional
density $\pi(x_i \mid \vec{x_{i-1}})$, which we used to inform the importance
distribution. In \gls{SMC}, the relationship between subsequent distributions
is less explicit, but it is assumed that they are related closely enough that
an importance distribution for $\pi_i$ can be easily transformed into one for
$\pi_{i+1}$. In particular, the sequence of importance distributions $\eta_i$
is constructed as
\begin{align}
  \label{eq:impint}
  \eta_i(x) = \int \eta_{i-1}(\xi) K_i(\xi, x) \d\xi,
\end{align}
where $K_i$ is a Markov kernel and the integral is over the space on which the
$\pi_i$ are defined. The choice of $K_i$ should be based on the percieved
relationship between $\pi_{i-1}$ and $\pi_i$. \textcite{del2006sequential}
propose the use of a \gls{MCMC} kernel with equilibrium distribution $\pi_i$.
That is,
\[
  K_i(\xi, x) = \max\left(1, \frac{q(x, \xi)\pi_i(\xi)}{q(\xi, x)\pi_i(x)}\right),
\]
where $q(\xi, x)$ is a proposal function such as a Gaussian distribution
centered at $\xi$ (see \cref{subsec:mfit}).

Although this method of building up $\eta$ appears straightforward, the
drawback is that the importance distribution itself becomes intractible. In
particular, evaluating $\eta_i(x)$ involves a $i$-dimensional integral of the
type in \cref{eq:impint}. As it is necessary to evaluate $\eta(x)$ pointwise to
perform \gls{IS}, this construction appears to have defeated the purpose of
providing an importance distribution for each $\pi_i$.
\textcite{del2006sequential} overcome this problem with two ``artificial''
objects. First, they propose the existence of \textit{backward} Markov kernels
$L_{i-1}(x_i, x_{i-1})$. For now, these kernels are arbitrary, and will be
precisely defined on a problem-specific basis. Second, they define an
alternative sequence of target distributions
\[
  \tilde{\pi}_i(\vec{x_i}) = \pi_i(x_i) \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)
\]
of increasing dimension. This brings us back to the setting described above in
\cref{subsec:sis}, namely of building up an importance distribution of
dimension $d$ sequentially through lower-dimensional distributions. We can
write $\tilde{\pi}_i$ in terms of $\tilde{\pi}_{i-1}$ by noticing that
\begin{align*}
  \frac{\tilde{\pi}_i(\vec{x_i})}{\tilde{\pi}_{i-1}(\vec{x_{i-1}})} 
  = \frac{\pi_i(x_i) \prod_{k=1}^{i-1} L(x_{k+1}, x_k)}
  {\pi_{i-1}(x_{i-1}) \prod_{k=1}^{i-2} L(x_{k+1}, x_k)}
  = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})},
\end{align*}
and hence
\[
  \tilde{\pi}_i = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})} \cdot \tilde{\pi}_{i-1}.
\]
Therefore, the importance weights for these new targets are defined recursively as
\begin{align}
  w(\vec{x_i}) 
    &= \frac{\tilde{\pi}_i(\vec{x_i})}{\eta_i(\vec{x_i})} \\
    &= \frac{\tilde{\pi}_{i-1}(\vec{x_{i-1}}) \pi_i(x_i) L(x_i, x_{i-1})}
           {\eta_{i-1}(\vec{x_{i-1}}) \pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
    &= w(\vec{x_{i-1}}) \cdot
      \frac{\pi_i(x_i) L_{i-1}(x_i, x_{i-1})}
           {\pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
    &\propto w(\vec{x_{i-1}}) \cdot
      \frac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}
           {\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}.
    \label{eq:smcwt}
\end{align}
The final key piece of information is to notice that, because the $L_i$ are
Markov kernels, $\pi_i$ is simply the marginal in $\vec{x_{i-1}}$ of
$\tilde{\pi}$. Therefore, a sample from $\tilde{\pi}_i$ automatically gets us a
sample from $\pi_i$, by considering only the $i$th component of $\vec{x_i}$.
These are all the ingredients we need to apply \gls{SMC}. The sequences of
kernels $L$ and $K$ should be chosen based on the problem at hand to inimize
the variance in the importance weights as well as possible. For a fixed choice
of $K_i$, the backward kernels $L_i$ which minimize this variance are called
the \defn{optimal} backward kernels. The full \gls{SMC} sampler algorithm is
presented as \cref{alg:smcsamp}.

\begin{algorithm}
  \caption{Sequential Monte Carlo sampler of \textcite{del2006sequential}.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\eta_1(x_1)$
      \Comment{Initialize the $k$th particle}
      \State $w^{(k)} \gets \dfrac{\gamma_1\left(x_1^{(k)}\right)}{\eta_1\left(x_1^{(k)}\right)}$
      \State Normalize the weights so that $\sum w = 1$
    \EndFor
    \For {$i = 2$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $K(x_{i-1}^{(k)}, x_i)$
        \Comment Extend the $k$th particle
        \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
      \If {$\ESS(w) < T$}
        \State Resample the particles according to $w$
        \For {$k = 1$ to $n$}
          \State $w^{(k)} \gets 1/n$
        \EndFor
      \EndIf
      \State Sample the $i$th component of $n$ particles with probabilities $w$
    \EndFor
  \end{algorithmic}
  \label{alg:smcsamp}
\end{algorithm}

@@

Ordinarily, Bayesian inference targets the posterior distribution $\Pr(\theta
\mid \vec{y})$. That is, in the Bayesian framework, model parameters with
higher posterior density are ``better'' in the sense that they offer a more
credible explanation for the observed data. Approximate Bayesian computation
offers an alternative metric for parameter credibility, namely the similarity
of simulated datasets to the observed data. If datasets simulated under the
model closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, suppose we have a distance measure $\rho$ defined on the space of all
possible data our model could generate. \gls{ABC} aims to sample from the joint
posterior distribution of model parameters and simulated datasets $\vec{z}$
which are within some small distance $\varepsilon$ of the observed data
$\vec{y}$,
\[
  \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon) =
  \frac{\Pr(\theta) \Pr(\vec{z} \mid \theta) \I_{A_{\varepsilon, \vec{y}}} (\vec{z})}
  {\int_{A_{\varepsilon, \vec{y}} \times \Theta} \Pr(\theta) \Pr(\vec{z} \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, \vec{y}}$ is an $\varepsilon$-ball around $\vec{y}$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
see in the next section, this distribution can be sampled from exactly. The
word ``approximate'' derives from the assumption that, for a suitably chosen
distance $\rho$ and a small enough $\varepsilon$, the marginal in $\vec{z}$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon) \d\vec{z}
  \approx \Pr(\theta \mid \vec{y}).
\]
To emphasize that the distribution on the left only approximates the posterior,
we will refer to it as the \textit{\gls{ABC} target distribution}. Note that in
many formulations, the distance function $\rho$ is defined as $\rho(S(\cdot),
S(\cdot))$ where $S$ is a function which maps data points into a vector of
summary statistics. This can be useful if the data are high-dimensional or of a
complex type, but it is not strictly necessary. For instance, if the data are
numeric and of low dimension, the distance function may simply be the Euclidian
distance~\autocite{sisson2007sequential}. For more complex data,
\textcite{nakagome2013kernel} proposed the use of a kernel function (defined in
\cref{subsec:treeshape}), an approach they dubbed \defn{kernel-\gls{ABC}}.

@@

Algorithms for performing \gls{ABC} fall into one of three categories:
rejection, \gls{MCMC}, and \gls{SMC}. To simplify the math, we shall restrict
the descriptions of these algorithms to the case of one simulated dataset per
parameter particle (the meaning of this will become clear shortly). The
extension to multiple datasets per particle will be given at the end.

Rejection ABC is the simplest method, and also the one which was first
proposed~\autocite{rubin1984bayesianly, tavare1997inferring}. The algorithm,
outlined in \cref{alg:abcrej}, repeats the following steps until a desired
number of samples from the target distribution are obtained. Parameter values
$\theta$ are sampled according to the prior distribution $\Pr(\theta)$. Then, a
simulated dataset $\vec{z}$ is generated from the model with the sampled
parameter values. By definition, the probability density of obtaining the
particular dataset $\vec{z}$ is $\Pr(\vec{z} \mid \theta)$. Finally, the
parameters are sampled if the distance of $\vec{z}$ from the observed data
$\vec{y}$ is less than $\varepsilon$ - that is, with probability
$\I_{A_{\varepsilon, \vec{y}}}(\vec{z})$. Putting this all together, the
parameters $\theta$ are sampled with probability proportional to
\[
  \Pr(\theta) \Pr(\vec{z} \mid \theta) \I_{A_{\varepsilon, \vec{y}}}(\vec{z}),
\]
which is exactly the numerator of the \gls{ABC} target distribution. Thus,
$\theta$ represents an unbiased sample from the approximate posterior.

@@

The most recently developed class of algorithm for \gls{ABC} is
\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, beaumont2009adaptive}. As
with \gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of
an existing Bayesian inference method, in this case the \gls{SMC} sampler
(\cref{subsec:smcsamp}). The sequence of target distributions is defined as
$\pi_i = \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon_i)$ for a decreasing
sequence of tolerances $\varepsilon_i$. The intention is for the algorithm to
progress smoothly through a sequence of target distributions which ends at the
\gls{ABC} approximation to the posterior. As discussed in
\cref{subsec:smcsamp}, the choices of the kernels $K$ and $L$ is
problem-specific, and so appropriate kernels must be chosen for \gls{ABC}.
Several options have been proposed~\autocite{beaumont2009adaptive,
sisson2007sequential, del2012adaptive}.

All the algorithms discussed in this section can be straightforwardly extended
to sample from the joint distribution
\[
  \pi_\varepsilon(\theta, z_1, \ldots, z_M \mid y),
\]
which is equivalent to associating $M$ simulated datasets to each parameter
particle instead of just one. The simulated dataset $z$ is replaced by
$\vec{z} = z_1, \ldots, z_M$, and the indicator function for the
$\varepsilon$-ball around $y$ is replaced by
\[
  \sum_{k=1}^M \I_{A_{\varepsilon, y}} (z_i).
\]
For \gls{ABC}-\gls{MCMC} and \gls{ABC}-\gls{SMC}, the proposal distribution
$q(\theta, \theta') \Pr(z \mid \theta')$ is replaced by
\[
  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i \mid \theta').
\]
